---
title: "CHIS"
author: "Aaron Lopez"
date: "4/29/2017"
output: html_document
---

### Load libraries
```{r, message=FALSE, warning=FALSE}

library(ggplot2)
library(reshape2)
library(dplyr)
library(ggthemes)
```

### Create a mosaicGG function to use on data sets going forward. In the function we are created a Chi-sq test to show the proportion of each group that is over or under represented in our data set. Those results are then used to create our heatmap plot. 

```{r}
mosaicGG <- function(data, X, FILL) {
  
  # Proportions in raw data
  DF <- as.data.frame.matrix(table(data[[X]], data[[FILL]]))
  DF$groupSum <- rowSums(DF)
  DF$xmax <- cumsum(DF$groupSum)
  DF$xmin <- DF$xmax - DF$groupSum
  DF$X <- row.names(DF)
  DF$groupSum <- NULL
  DF_melted <- melt(DF, id = c("X", "xmin", "xmax"), variable.name = "FILL")
  DF_melted <- DF_melted %>% 
    group_by(X) %>% 
    mutate(ymax = cumsum(value/sum(value)),
           ymin = ymax - value/sum(value))
  
  # Chi-sq test
  results <- chisq.test(table(data[[FILL]], data[[X]])) # fill and then x
  resid <- melt(results$residuals)
  names(resid) <- c("FILL", "X", "residual")

  # Merge data
  DF_all <- merge(DF_melted, resid)
  
  # Positions for labels
  DF_all$xtext <- DF_all$xmin + (DF_all$xmax - DF_all$xmin)/2
  index <- DF_all$xmax == max(DF_all$xmax)
  DF_all$ytext <- DF_all$ymin[index] + (DF_all$ymax[index] - DF_all$ymin[index])/2
  
  # plot:
  g <- ggplot(DF_all, aes(ymin = ymin,  ymax = ymax, xmin = xmin, 
                          xmax = xmax, fill = residual)) + 
  geom_rect(col = "white") +
  geom_text(aes(x = xtext, label = X),
            y = 1, size = 3, angle = 90, hjust = 1, show.legend = FALSE) +
  geom_text(aes(x = max(xmax),  y = ytext, label = FILL),
            size = 3, hjust = 1, show.legend = FALSE) +
  scale_fill_gradient2("Residuals") +
  scale_x_continuous("Individuals", expand = c(0,0)) +
  scale_y_continuous("Proportion", expand = c(0,0)) +
  theme_tufte() +
  theme(legend.position = "bottom")
  print(g)
}
```

### Use mosaicGG function on some data sets to see what we get. I took out adult from the exercise becuase the data set was unavailable. I added an example using the diamonds data set since that is part of the ggplot package.

```{r, message=FALSE, warning=FALSE}
# mtcars: am described by cyl
mosaicGG(mtcars, "cyl", "am")

# diamonds: clarity described by color
mosaicGG(diamonds, "color", "clarity")

# Vocab: vocabulary described by education
library(car)
mosaicGG(Vocab, "education", "vocabulary")
```

