---
title: "Predicting San Jose Home Prices"
author: "Aaron Lopez"
date: "6/11/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Description

Affordable housing is a major concern in the Bay Area of California. Many people are priced out of affordable housing options where the median home price has reached $1 million in San Jose, California.^1^ This trend impacts low and median income families looking to secure home ownership. Politicians and voters have voiced their concerns and willingness to address this issue by approving new policies at the ballot box. Last November voters passed Measure A allowing the city to issue \$950 million in new bonds for affordable housing.^1^ The current mayor of San Jose, Sam Liccardo, has said, "San Jose is facing an affordable housing crises."^2^ To help address this issue I'd like to take a look at recent median home price trends and also predict how prices will change in the near term. This information will help to keep track of affordability and give insights into the effectiveness of government policy changes.
  
There are two major client groups. First, the city of San Jose who has made housing affordability a priority as the city continues to grow and city officials create urban planning policy. Politicians have expressed their desire to introduce an impliment policy to make housing more affordable for low and middle income residents. Understanding the trend and where housing prices are headed will help politicians determine what policy may be needed or developed. The second group of clients would be low and median income individuals seeking to find affordable housing and deciding when and if San Jose home price are affordable today or in the near future. This information can help them make an infored decision on when and if to purchase a home in the near future. 

To predict near term median home prices I will use three different model approaches, linear regression, time-series (ARIMA), and regression tree. I will then compare each model to actual median home prices from a test data set. Data will be used from the city of San Jose outlined below.
  
## Data Set

The main source of data will be from the City of San Jose <http://data.sanjoseca.gov/home>. Economic and labor data including unemployment, total jobs, jobs by sector, median home prices, median rental prices will all be utilized along with broader economic data such as average mortgage rates from FRED <https://fred.stlouisfed.org/graph/?g=NUh#0>. The data set is approximately 6 years of monthly data from 2009-2015 after merging and cleaning up any missing values. There are 6 csv files all under the "Data" file. 

apt_rents.csv
home_prices.csv
jobs_by_sector.csv
mortgagerates.csv
total_jobs.csv
unemployment.csv

San Jose provides relatively clean uniform data however it can be a little outdated and incomplete. I would have like to include crime data but this was missing at the time of my analysis and a time frame for requesting the data was uncertain. Including this information may be useful in further analyizing median home price trends.

## Data Wrangling

The main steps to cleaning up the data were simplifying the variable names, formatting strings, and converting objects. The city of San Jose has some consistency in their data sets but to combine all the data I had to do some clean up of individual csv files before combining it all into one data frame for analysis. 

```{r message=FALSE}
# Load Data and Libraries

library(dplyr)
library(tidyr)
library(reshape2)
library(plyr)
jobs_by_sector  <- read.csv("Data/jobs_by_sector.csv")
home_prices     <- read.csv("Data/home_prices.csv")
total_jobs      <- read.csv("Data/total_jobs.csv")
unemployment    <- read.csv("Data/unemployment.csv")
apt_rents       <- read.csv("Data/apt_rents.csv")
mortgagerates   <- read.csv("Data/mortgagerates.csv")
jobs_by_sector  <- tbl_df(jobs_by_sector)
total_jobs      <- tbl_df(total_jobs)
unemployment    <- tbl_df(unemployment)
apt_rents       <- tbl_df(apt_rents)
mortgagerates   <- tbl_df(mortgagerates)

# Bring column names (dates) into a unique column so table is in long format
jobs_by_sector <- melt(jobs_by_sector, variable.name = "Sector")

# Rename column to correct variable name
colnames(jobs_by_sector)[2] <- "Date"

# Move sector names into column names moving the table back to wide format
jobs_by_sector <- spread(jobs_by_sector, Sector, value)

# Rename column names to be short and concise
colnames(jobs_by_sector)[3] <- "Education"
colnames(jobs_by_sector)[4] <- "Finance"
colnames(jobs_by_sector)[6] <- "Hospitality"
colnames(jobs_by_sector)[8] <- "Mining"
colnames(jobs_by_sector)[9] <- "Other"
colnames(jobs_by_sector)[10] <- "Business"
colnames(jobs_by_sector)[11] <- "Public"
colnames(jobs_by_sector)[12] <- "Trade"

# Changing dates from string to date format. as.Date returns NA for "Sept" 
# abbreviation so first we'll change that to "Sep"
jobs_by_sector$Date <- gsub("Sept", "Sep", jobs_by_sector$Date)
# Adding a day
jobs_by_sector$Date <- paste0("01.", jobs_by_sector$Date)
# Format date 
jobs_by_sector$Date <- as.Date(jobs_by_sector$Date, format = "%d.%b.%Y", "%b/%d/%Y")

# Setup column names for gather function to put back into long format
jobs_by_sector <- jobs_by_sector %>% 
  gather(Construction, Education, Finance, Information, Hospitality, 
         Manufacturing, Mining, Other, Business, Public, Trade, Unclassified, 
         key = "JobSector", value = "NumJobs")

head(jobs_by_sector)

# Setting up dates "Sept" replaced with "Sep"
home_prices$Date <- gsub("Sept", "Sep", home_prices$Date)

# Adding a day then format to date
home_prices$Date <- home_prices$Date %>% 
    paste("01", sep = " ") %>% 
    as.Date(format = "%b %Y %d", "%b/%d/%Y")

# Same steps for total_jobs df
total_jobs$Date <- gsub("Sept", "Sep", total_jobs$Date)
total_jobs$Date <- total_jobs$Date %>%
    paste("01", sep = " ") %>% 
    as.Date(format = "%b %Y %d", "%b/%d/%Y")

# Again for unemployment df
unemployment$Date <- gsub("Sept", "Sep", unemployment$Date)
unemployment$Date <- unemployment$Date %>%
    paste("01", sep = " ") %>%
    as.Date(format = "%b %Y %d", "%b/%d/%Y")

# Slightly different strategy for apt_rents df. Dates are in integer format. 
# First I added the day then converted to date format.
apt_rents$Date <- paste0("01/", apt_rents$Date)
apt_rents$Date <- apt_rents$Date %>% 
    as.Date(format = "%d/%m/%Y", "%b/%d/%Y")

# Cleaninging up mortgage rates with more consies variable names then date format
colnames(mortgagerates)[1] <- "Date"
colnames(mortgagerates)[2] <- "Rates"
mortgagerates$Date <- mortgagerates$Date %>% 
    as.Date(format = "%Y-%m-%d", "%b/%d/%Y")

# Combine all data frames by date nd arranging in desending order
clean_df <- join_all(list(jobs_by_sector,apt_rents,home_prices,total_jobs,
                          unemployment, mortgagerates), by="Date", type="left")
clean_df <- arrange(clean_df, desc(Date))

# Cleaning up some formating to convert data types to numeric by removing dollar 
# signs and commas
clean_df$Condo.Townhome <- gsub("\\$|,", "", clean_df$Condo.Townhome)
clean_df$Single.Family.Home <- gsub("\\$|,", "", clean_df$Single.Family.Home)
clean_df$X2.bedroom <- gsub("*,", "", clean_df$X2.bedroom)
clean_df$X1.Bedroom <- gsub("*,", "", clean_df$X1.Bedroom)
clean_df$Average <- gsub("*,", "", clean_df$Average)

# Converting integers to numeric across variables
clean_df[3:9] <- lapply(clean_df[3:9], as.numeric)

# Converting job secotrs from characters to factors
clean_df$JobSector <- as.factor(clean_df$JobSector)

# Renaming variables with more descrptive concise names
colnames(clean_df)[4] <- "AvgAptRent"
colnames(clean_df)[5] <- "Avg1bdAptRent"
colnames(clean_df)[6] <- "Avg2bdAptRent"
colnames(clean_df)[7] <- "AvgPriceHome"
colnames(clean_df)[8] <- "AvgPriceCondo"
colnames(clean_df)[9] <- "TotalJobs"
colnames(clean_df)[10] <- "URateSJ"
colnames(clean_df)[11] <- "URateSJMetro"

# removing missing data
clean_df <- na.omit(clean_df)

# adding a month variable for exploritory analyis
month_prices <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
clean_df$Month <- factor(format(clean_df$Date, "%b"), levels = month_prices)

# Create final new csv file
write.csv(clean_df, file = "Data/clean_df.csv")

# Summary of the final data frame
str(clean_df)
```

## Exploratory Analysis

```{r, echo=FALSE}
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)

```

To start exploring the data set let's look at correlations between the variables. Rental prices, house prices, condo prices, jobs, and umeployment all are highly corrilated either positively or negatively as one would expect. The two variables that are the most interesting are rates and number of jobs. Rates refers to the average 30-year mortgage interest rate and number of jobs is tied to each job sector i.e. construction, business, finance. The positive or negative correlations are what you would expect but not as strong as some of the other variables. 

```{r, echo=FALSE, results='asis'}
library(knitr)
# Looking at correlations 
kable(cor(clean_df[3:12], use="complete.obs"))
```

```{r, echo=FALSE, message=FALSE}
# Time series of avergae home prices
ggplot(clean_df, aes(Date, AvgPriceHome/1000)) + geom_point() + geom_smooth() + labs(title = "Price of Average Single Family Home", y = "Price ($1,000s)") + theme_economist()
```
Next let's start visualizing some of the data with a few plots. First, let's look at a time series of home prices. The data starts after the bottom of the financial crises in 2009 and then there is a second drop in prices a few years after with a subsequent rebound in more recent years to pre-recession levels. There seems to be some short term trends or seasonality in housing prices that may be easier to visualize with other plots but overall the data set reflects a time of home price appreciation in San Jose.

```{r, echo=FALSE, message=FALSE}
# Time series of average home prices line, drop off at the end of each year. Best time to buy a home?
ggplot(clean_df, aes(Date, AvgPriceHome/1000)) + geom_line() + labs(title = "Price of Average Single Family Home", y = "Price ($1,000s)") + theme_economist()
```
With a time series in a simple line graph its easier to see the up and down movement of home prices that perhaps reveals a seasonal trend. I would expect home prices to rise during the summer buying months when people are perhaps more inclined to go out and look for homes. During the winter months people tend to stay in more although winters are not that harsh in San Jose. A box plot would be useful to analyze this on a monthly basis. 

```{r, echo=FALSE, message=FALSE}
ggplot(clean_df, aes(Month, AvgPriceHome/1000, group=Month, fill=Month)) + geom_boxplot() + theme_economist() + theme(legend.position="none") + labs(title = "Single Family Home Box Plot", y = "Price ($1,000s)")
```
From the box plot above we can see home prices tend to be higher during the summer months of April, May, June, and July then start to come down as the winter months begin. From this data January and February seems to be the best time to buy if you're looking for a new home.

```{r, echo=FALSE, message=FALSE}
# Time series of jobs by sector 2008-2016
ggplot(clean_df, aes(Date, NumJobs, col = JobSector)) + geom_line() + labs(y = "Number of Jobs") + theme_economist()
```
The second set of time series plots shows the total number of jobs by job sector. Education, Business, and Trade are the top sectors in terms of number of jobs. The education sector shows high yearly variation most likely due to teachers getting let go for the summer months when students aren't in school.

```{r, echo=FALSE, message=FALSE}
# Box plot of jobs by sector
ggplot(data=clean_df, aes(x=JobSector, y=NumJobs, group=JobSector, fill=JobSector)) + geom_boxplot() + coord_flip() + guides(fill=FALSE) + theme_economist() + labs(x = "Job Sector", y = "Number of Jobs", title = "Job Sector Box Plot")
```
If we look at a box plot the top jobs sectors also have the most variabliility in the number of jobs. Manufacturing is the fourth place sector but shows a lot less variablity when compared to the other top sectors. It's importatnt to consider the data set is mostly in a time of strong economic growth so manufacturing may or may not be more variable through a full business cycle that includes both weak and strong economic growth.

```{r, echo=FALSE, message=FALSE}
# Time series of Education sector jobs with visable seasonality
ggplot(subset(clean_df, JobSector=="Education")) + geom_line(aes(Date, NumJobs, col = JobSector)) + theme_economist() + labs(title = "Education Jobs Sector", y = "Number of Jobs") + theme(legend.position="none")
```
Lastly, I've pulled out the Education sector to show how the number of education jobs drops siginifcantly each year during the summer months as you might expect. It's also interesting how the variation seems to be in the same range from 2009 - 2012 then spikes to a new level in 2013. Perhaps more teachers were hired overall due to a budget increase or the methodogy changed for counting education jobs. 

```{r, echo=FALSE, message=FALSE}
# Histogram of San Jose unemployment rate
ggplot(clean_df, aes(URateSJ, fill = cut(URateSJ, 25))) + geom_histogram(show.legend = FALSE, bins = 25) + scale_fill_discrete(h = c(240, 100)) + theme_economist() + labs(x = "Unemployment Rate (%)", y = "Number of months", title = "Uneployment Rate Histogram")
```
Taking a look at the unemployment rate in our data set I thought a histogram might be intersting to look at. When counting the number of observations for each unemployment rate in a historgram we can see the rate stood the most months at 5.4 and 5.5% and ranged from 4.6 to 12.6%. 

```{r, echo=FALSE, message=FALSE}
ggplot(clean_df, aes(Date, URateSJ)) + geom_line() + theme_economist() + labs(title = "San Jose Unemployment Rate", y = "Unemployment Rate (%)")
```
The umeployment rate has trended downward over the range of our data set. This would indicate we are looking only at a time when economic conditions were improving and not a full market cycle that would reflect both improving and declining economic conditions. 

```{r, echo=FALSE, message=FALSE}
# Time series of average 30-year mortage rates 
ggplot(clean_df, aes(Date, Rates)) + geom_point() + geom_smooth() + theme_economist() + labs(y = "Average 30-year Mortgage Rate (%)")
```
Lastly, a final time series plotting average 30-year mortgage rates shows a historical fall in rates since 2008 that bottomed in late 2012 and have remained range bound between 3.5 and 4.5% since. This may have helped the rise in average home prices recently as lower rates create a more favorable lending environment for borrowers to purchase homes. Rates started to trend downward due to the financial crises and have remained historically low ever since. 

## Model Building

## Results and Discussion

## Appendix

#### References
1. <https://www.nar.realtor/sites/default/files/reports/2017/embargoes/2017-q1-metro-home-prices/metro-home-prices-q1-2017-single-family-2017-05-15.pdf>
2. <http://yesonaffordablehousing.org/>
3. <http://www.sanjoseca.gov/index.aspx?NID=5256>