---
title: "Statistical Analysis Report"
author: "Aaron Lopez"
date: "5/11/2017"
output: html_document
---

```{r setup, include=FALSE}
```

### Statistical Analysis Report for Ch.5

The first time series of home prices (AvgHomePrice.png) shows the slide of home prices during the financial crises. Some observations are the second drop in prices a few years after the financial crises and the subsequent rebound in more recent years to pre-recession levels. There also seems to be a annual seasonal trend with prices rising in the beginning of the year then falling again at the end of the year. (AvgHomePriceLine.png)

The second set of time series shows total number of jobs by sector (JobSector.png). Education, Business, and Trade are the top sectors in terms of number of jobs. Education also shows strong seasonal trends with summer months as school is not in session and there are less teaching jobs (EducationSector.png)

The third set of plots relate to total jobs and the unemployment rate (UnempRateJob.png) there is a clear negative correlation between the total number of jobs and the unemployment rate. As the unemployment rate increases the total number of jobs decreases. When counting the number of observations for each unemployment rate we can see the rate stood the most months at 5.4 and 5.5% and ranged from 4.6 to 12.6 (URateSJ.png)

Lastly, a final time series plotting average 30-year mortgage rates shows a historical fall in rates since 2008 that bottomed in late 2012 and have remained range bound between 3.5 and 4.5% since. This may have helped the rise in average home prices recently as lower rates create a more favorable lending environment for borrowers to purchase homes (Rates.png)

```{r}
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)


# Time series of avergae home prices
ggplot(clean_df, aes(Date, AvgPriceHome/1000)) + geom_point() + geom_smooth()
# Time series of average home prices line, drop off at the end of each year. Best time to buy a home?
ggplot(clean_df, aes(Date, AvgPriceHome)) + geom_line()

# Time series of jobs by sector 2008-2016
 ggplot(clean_df, aes(Date, NumJobs, col = JobSector)) + geom_point()
# Time series of Education sector jobs with visable seasonality
 ggplot(subset(clean_df, JobSector=="Education")) + geom_point(aes(Date, NumJobs, col = JobSector))

# Total jobs and unemployment rate
ggplot(clean_df, aes(URateSJ, TotalJobs)) + geom_point()
# Histogram of San Jose unemployment rate
ggplot(clean_df, aes(URateSJ, fill = cut(URateSJ, 25))) + geom_histogram(show.legend = FALSE, bins = 25) + scale_fill_discrete(h = c(240, 100))

# Time series of average 30-year mortage rates 
ggplot(clean_df, aes(Date, Rates)) + geom_point() + geom_smooth()

```

